---
title: "NYPD Shooting Incident (Historic)"
output:
  pdf_document: default
  html_document: default
date: '2022-07-14'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load_packages, message=FALSE}
library(tidyverse)
library(lubridate)
```

```{r read_in_data, message=FALSE}
data <- read_csv('https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv')
```
The source data is provided by the New York Police Department listing shooting
incidents from 2006 to 2021. The data set includes information on the the perpetrator, 
the victim, the location of the incident and when the incident took place.

# The Question

After the a brief examination of the data, a question that comes to mind is:

Which boroughs in New York are safe to live in?

It is important to note that the data used is historic data of shooting incidents
in New York from 2006 to 2021. Therefore, "safety" from the question is going
to be defined as the chance of not getting involved in a shooting incident and 
dying.

With that out the way, it's time to start tidying up the data set.

# Data Cleaning
```{r tidy_data}
# took out some unnecessary columns
# and changed the date column to a date data type
data <- data %>% 
  select(c(OCCUR_DATE, BORO, STATISTICAL_MURDER_FLAG, VIC_AGE_GROUP,
           VIC_SEX, VIC_RACE)) %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE))

# made some categorical columns into factors
cols_to_factor <- c('BORO', 'VIC_AGE_GROUP', 'VIC_SEX', 'VIC_RACE')
data <- data %>%
  mutate_at(cols_to_factor, factor)

# No NA values seen for these columns
summary(data)

head(data)
```

# Visualization and Analysis

To start looking at safety, I want to look at the chance that a person may die 
from a shooting incident from any borough.

## Chance of Death

To find the likelihood that a person may die from a shooting, I decided to 
look at the column named 'STATISTICAL_MURDER_FLAG'. According to NYPD who 
provided this data set, this column is a conditional column, where TRUE would 
indicate a victim's death resulting from a shooting incident, i.e. a murder. 

The task is then simple, to find the chance of death from getting involved in 
any shooting incident in New York would then be the sum of the TRUE values in 
divided by total incidents.

```{r death_chance}
# grouping by borough and calculating the chance of death from 
# dividing the TRUE values from total incidents of each group.
boro <- data %>%
  group_by(BORO) %>%
  summarize(murders = sum(STATISTICAL_MURDER_FLAG) / n())

# creating bar graph
boro %>%
  ggplot(aes(x = BORO, y = murders, fill = BORO)) +
  geom_bar(stat = 'identity', width = 1, color = 'white') + 
  labs(title = "Chance of Death from Shooting Incidents By Borough", 
       y = 'Percentage', x = 'Borough', color = 'Boroughs') + 
  scale_fill_discrete(name=NULL)

```
It seems that all five boroughs in New York have roughly a 20% chance of murder
from any shooting incident. That's every 1 in 5 shooting incident the victim is 
likely to get killed. That does not sound very good.

Now let's talk about the victim.

## The Victim

To find the safety of the boroughs, we are going to look at the demographics of
the victims. More specifically, we are going to examine their age, race, and 
sex.

### Age


```{r victim_age, message=FALSE}
# group by victim age showing total for each age group
victim_ages <- data %>%
  group_by(BORO, VIC_AGE_GROUP) %>%
  summarize(num_cases = n())

# creating bar chart
victim_ages %>%
  ggplot(aes(x = BORO, y = num_cases, fill = VIC_AGE_GROUP)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(title = 'Age Group of Victims By Borough', 
       x = 'Borough', y = 'Number of Cases') + 
  scale_fill_discrete(name='Age Group')

```
From the chart displayed above, there's a few things that can be glimpsed. 
Initially, we spoke about the probability that the victim is killed from 
any shooting in each borough. Here we can see that Brooklyn has the most cases
and Bronx coming in close. Staten Island has the least amount of incidents 
in comparison. 

Next, we notice that there is a large discrepancy in the age groups of the
victim. For every borough, the vast majority of them were between ages 18-44, 
the age range for young adults. People falling into that age group would be at
a higher risk of being involved in shooting incidents, supposedly. The reason 
could be that that age group is where the most social interactions would occur. 


### Race


```{r victim_race, message=FALSE}
# group by borough and race of victims and get the total count for grouping
victim_race <- data %>%
  group_by(BORO, VIC_RACE) %>%
  summarize(num_cases = n())

# bar chart
victim_race %>%
  ggplot(aes(x = BORO, y = num_cases, fill = VIC_RACE)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(title = 'Racial Identity of Victims By Borough', 
       x = 'Borough', y = 'Number of Cases') + 
  scale_fill_discrete(name='Race') + 
  theme(axis.text.x = element_text(angle = 12))
```
The first thing that pops out from the chart is the green bars. The green bars
indicate victims that are considered Black. There is an overwhelming number of 
cases where Blacks end up as victims to shootings. White Hispanics are the 
second-most typical race to be victims but not by a long shot. Furthermore, at 
a quick glance, the distribution between races for each borough looks very 
similar to one another ranking:

1. Black
2. White Hispanic
3. Black Hispanic
4. White
5. Other 

People from the lower end of the socioeconomic spectrum is more likely to be 
victims of shootings whether it be by chance or involvement of some sort. 

### Sex

So far we have seen that victims tend to be Black young adults that are getting
involved in the shooting incidents in New York. From this we can sort of infer
that the sex of the victims are most likely going to be predominantly male. But 
to confirm this speculation, let's look at the data.

```{r victim_sex, message=FALSE}
# grouping by borough and sex identity of victims
victim_sex <- data %>%
  group_by(BORO, VIC_SEX) %>%
  summarize(num_cases = n())

# bar chart
victim_sex %>%
  ggplot(aes(x = BORO, y = num_cases, fill = VIC_SEX)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(title = 'Sex Identity of Victims By Borough', 
       x = 'Borough', y = 'Number of Cases') + 
  scale_fill_discrete(name='Sex')
```
We can see that the speculation that males would be the predominant sex for the 
victims was correct. We see that every borough is the same in that males are 
extremely more likely to be victims to shootings.


## Cases Over Time

So we have a general identification for the victims and we find that there have
been a significant amount of shooting cases in each borough. But have they 
changed over the years? Are shootings still rampant or have they decreased over
time? 

```{r time, message=FALSE}
# round to closest year
data$OCCUR_DATE <- floor_date(data$OCCUR_DATE, unit = 'year')

# group by year and borough
time_data <- data %>%
  group_by(BORO, OCCUR_DATE) %>%
  summarize(num_cases = n())

# line graph 
time_data %>%
  ggplot(aes(x = OCCUR_DATE, y = num_cases, color = BORO)) +
  geom_line(size = 1.5, alpha = 0.65) +
  geom_point(size = 2.5) + 
  labs(title = 'Yearly Cases By Borough', 
       x = 'Year', y = 'Number of Cases') + 
  scale_color_discrete(name = 'Borough')

```
Starting from 2006 when the data set started, there has been a general decrease
in shootings in every borough except Staten Island which remains relatively 
constant albeit their low number of cases. Interestingly, we see that shootings
suddenly spiked in 2020 and boroughs like the Bronx and Manhattan setting new
highs in the number of cases in 2021. 

Generally the cases were decreasing in areas where shootings were a serious 
problem. This may be due to some gun regulations over the years from the state
of New York or just in terms of education. Yet there was a spike in recent years. 
Coincidentally, 2020 to 2021 was the year COVID-19 emerged and was a major problem. 
Within that context, the spike in cases may be a result of several things such as
mental health issues or disputes from a loss of income source. 

# Modeling Predicted Murders

We are going to model predicted murders by the number of total shootings that 
occurred in New York.

```{r model, message=FALSE}
# creating a data set for total shootings and incidents considered murders
murders <- data%>%
  group_by(BORO) %>%
  summarize(total_murders = sum(STATISTICAL_MURDER_FLAG == TRUE),
            total_alive = sum(STATISTICAL_MURDER_FLAG == FALSE),
            total_cases = n()) %>%
  ungroup() %>%
  mutate(cum_shootings = cumsum(total_cases)) %>%
  mutate(cum_murders = cumsum(total_murders))

# creating linear model
model <- lm(cum_murders ~ cum_shootings, data = murders)

# creating prediction
pred <- murders %>%
  mutate(pred = predict(model))

# graphing line graph of model
pred %>%
  ggplot(aes(x = cum_shootings, y = cum_murders, color = 'blue')) +
  geom_point(alpha = 0.65, size = 1.5) +
  geom_line(aes(x = cum_shootings, y = pred, color = 'red'), alpha = 0.65, size = 1.5) + 
  labs(title = 'Model Predicting Amount Cases Considered Murders', 
       x = 'Number of Shootings', y = 'Number of Murders') + 
  scale_color_manual(labels = c('Actual', 'Predicted'), values = c('blue', 'red'))
```

The original data showed a relatively linear relationship once the cumulative sum
was taken to find the total shootings and total murder incidents. A linear model
was then used to try to predict the number of murders given a number of total shootings.
The model was shown to be very close to what was perceived in the original data, 
landing almost on top of each other. Looking at the graph, again it looks like
the number of murders compared to the total cases in around 1 in 5.

# Conclusion

Going back to the question: which boroughs in New York are safe to live in, the 
answer probably Staten Island. According to cases over time, they have a 
relatively low number of shooting incidents in comparison to the other boroughs
and did not have a heavy resurgence in recent years.

However, regardless where one might live, there is a 1 in 5 chance a shooting case
results in death of the victim. Victims tend to be young adults (ages 18-44) black males.
So if you don't belong to these demographics then the risk of danger should be reduced.

It is true that these characteristics indicate a high chance of becoming a victim
to shootings according to the data. However, we do not know whether there was a 
relationship between the perpetrator and victims nor the motive behind the incidents
according to the data provided. Therefore, we can't say exactly which borough is safer
because we do not know whether these incidents are personal or random occurrences. 
One would be better than the other simply by just not getting involved. 

Furthermore, we defined safety in this case to be not getting involved in a shooting
incident and dying however safety is more than just shootings. It would be interesting to see
what percent of criminal charges were shootings and their motives. Also it would 
be interesting to find why Brooklyn has such a high number of shooting incidents.

# Personal Bias

Coming into this data exploration, I may have had confirmation bias where I 
thought the victims to be mostly young black adults and I was tempted to look solely on
that data. In terms of bias management, I tried to look objectively at the data
and not try to overly emphasize any demographic to create a story that pointed 
heavily in one direction. I tried to make the story more like a layer of discovery
and included as much source of comparison as the data can provide to give a more
generalized picture.

# Session Info
```{r session_info, message=FALSE, echo=FALSE}
sessionInfo()
```